Linear stability of magnetized massive protoplanetary disks

### 2014-02-12 paper/results.tex
backup

\section{Results: MRI in self-gravitating disks}

@@

In this section we focus on the MRI. We consider polytropic disks,
which have a well-defined thickness, and a vertical field. 
We set the upper disk boundary at $Z_s=0.99H$ and fix $\beta=100$ and
$k_xH=0.1$. The numerical resolution is $N_z=256$.

@@

\subsection{Influence of self-gravity on the MRI through the linear
  response}

Our goal here is to examine whether or not self-gravity can amplify
the density perturbations caused by the MRI. We choose a massive 
isothermal disk with $Q=0.2$, which is still expected to be marginally
stable to gravitational instability \citep[][who find
a critical value of  $Q\simeq 0.2$]{mamat10}.  
The upper disk boundary is set to $Z_s=H$. The field is vertical.

@@

\subsubsection{Ideal case}
We first consider the limit of ideal MHD by adopting a uniform
resistivity with $\Lambda_0=100$. %Non-zero density perturbations
%require $k_x\neq0$ so we consider linear modes as a function of $k_x$.

@@

\subsubsection{Qualitative interpretation} 
In order to make sense of the above results, we consider ideal MHD and
a vertical field. For this discussion we will ignore stratification
and set $d/dz\to \imgi k_z$ where $k_z$ is a vertical 
wavenumber. Equivalently, we are considering regions close to the disk
midplane. The governing equations are then

\begin{align}
  &  0= v_A^2k^2\dvx + \imgi\sigma\left(\imgi\sigma \dvx - 2\Omega\dvy + \imgi k_x \w\right),\label{simp1}\\
  &  0= v_A^2k_z^2\left(\dvy + \frac{\imgi S
  }{\sigma}\dvx\right) + \imgi\sigma\left(\imgi\sigma\dvy +
  \frac{\kappa^2}{2\Omega}\dvx\right),\\
  & 0 = -k_z^2\w + \frac{\sigma^2}{c_s^2}W + \sigma k_x \dvx, \label{simp3}\\
  & 0 = k^2\dphi + \frac{\Omega^2}{c_s^2Q}W,\label{simp4}
\end{align}
where $k^2 = k_z^2 + k_x^2$.

@@

We now include self-gravity. In general, the MRI will have some
density perturbation so $W^{(0)}\neq0$. The Poisson equation implies
an associated potential perturbation,    
\begin{align} 
  \dphi = -\frac{\Omega^2}{c_s^2 Q k^2} W^{(0)}.
\end{align}
Physically, we expect $k^2\geq0$, so that a positive (negative) density
perturbation causes a negative (positive) potential perturbation. We
then regard $\dphi$ to alter the Cowling solution, so
that $\dvx^{0} \to \dvx^{0} + \dvx^{(1)}$ and similarly for $\dvy$ and
$W$. Then 
\begin{align}
  &   k_x\sigma \dphi = v_A^2k^2\dvx^{(1)} + \imgi\sigma\left[\imgi\sigma
  \dvx^{(1)} - 2\Omega\dvy^{(1)} + \imgi k_x W^{(1)}\right], \label{simp_pert1}\\ 
  &  0= v_A^2k_z^2\left[\dvy^{(1)} + \frac{\imgi S
    }{\sigma}\dvx^{(1)}\right] + \imgi\sigma\left[\imgi\sigma\dvy^{(1)} +
  \frac{\kappa^2}{2\Omega}\dvx^{(1)}\right],\label{simp_pert2}\\
  & k_z^2\dphi  = \left(\frac{\sigma^2}{c_s^2}-k_z^2\right)W^{(1)} +
  \sigma k_x \dvx^{(1)}.\label{simp_pert3} 
\end{align}
Now, if the perturbations to the magnetic field remain
unchanged, i.e. the mode remains close to the standard MRI, then
$\dvx^{(1)} \sim 0$ and $\dvy^{(1)}\sim0$, so Eq. \ref{simp_pert2} is
satisfied. Eq. \ref{simp_pert1} then require $\dphi +
W^{(1)}\sim0$. This is compatible with Eq. \ref{simp_pert3} if 
\begin{align}
  \left|k_z^2\right| \gg \left|\frac{\sigma^2}{c_s^2}\right|. \label{cond}
\end{align}
These assumptions imply
\begin{align}
  W^{(1)} \sim \frac{\Omega^2}{c_s^2 Q k^2} W^{(0)},\label{feedback}
\end{align} 
which indicates a non-zero density perturbation due to the
MRI can be amplified by self-gravity. 
%However, the magnetic field
%perturbations remain that in the

By introducing $\hat{k}_z=k_zH$, Eq. \ref{cond} can be
non-dimensionalized to 
\begin{align}
  \left|\frac{\hat{\sigma}^2}{f^2\hat{c}_s^2\hat{k}_z^2}\right|\ll 1.  
\end{align}

%\begin{align}
%  
%\end{align}
%However, in the full problem $\dphi$ appears in the $\mathcal{G}$

%We write the horizontal velocities and enthalpy perturbations as
%\begin{align}
%  \dvx &= u + \Delta u,\\
%  \dvy &= v + \Delta v,\\
%  W    &= w + \Delta w,
%\end{align}
%where $u$, $v$ and $w$ is the solution to
%Eq. \ref{simp1}---\ref{simp4} in the Cowling approximation, in which
%the Poisson equation is ignored and the potential perturbation set to
%zero.

>>>>

\section{Results: MRI in self-gravitating disks}
In this section we focus on the MRI. We use the vertical field setup
of case 1. The numerical resolution is $N_z=256$.

@@

Here we consider polytropic disks, which have a well-defined
thickness. Numerically, the upper disk boundary is set to
$Z_s=0.99H$. We fix $\beta=100$ and $k_xH=0.1$.

@@

\subsection{Influence of self-gravity on the MRI through the linear
  response}  
Our goal here is to examine whether or not self-gravity can amplify
the density perturbations caused by the MRI. We choose a massive 
isothermal disk with $Q=0.2$, which is still expected to be marginally
stable to gravitational instability \citep[][who find
a critical value of  $Q\simeq 0.2$]{mamat10}.   
The upper disk boundary is set to $Z_s=H$.

@@

\subsubsection{Ideal case}
We first consider the limit of ideal MHD by adopting a uniform
resistivity with $\Lambda_0=100$. %Non-zero density perturbations
%require $k_x\neq0$ so we consider linear modes as a function of $k_x$.   
Fig. \ref{gravity_energy} shows the maximum gravitational potential
energy perturbation, as a function of $k_x$ for several values of
$\beta$. The potential perturbation is negligible for all cases when 
$k_xH\lesssim 0.5$, since the MRI becomes incompressible as $k_x\to
0$.

@@

For $\beta\gg 1$, i.e. a weak field, density perturbations are
negligible and the standard MRI operates. However, as $\beta$ is
lowered we find non-negligible potential perturbation for
$k_xH=O(1)$. This suggets that in a strongly magnetized disk that
still permits the MRI, the associated density perturbation can be
important when the disk is self-gravitating.

Fig. \ref{mri_massive} shows an example of a MRI mode with
non-negligible density perturbation. The growth rate $\gamma\simeq
0.4\Omega$ is comparable to that of the most unstable mode
($\gamma\simeq 0.6\Omega$ for $k_x\simeq0$). 
We therefore expect significant density perturbations to grow
dynamically, even though the disk is expected to be gravitationally
stable.

%dynamical
%timescales 
%($\gamma\sim0.4\Omega$)

\begin{figure}
  \includegraphics[width=\linewidth]{figures/result_ideal_sg}
  \caption{Example of a linear MRI mode with non-zero density and
    potential perturbation, in a strongly magnetized massive disk
    ($\beta=10$ and $Q=0.2$) in the limit of ideal MHD
    ($\Lambda_0=10^2$).   
    \label{mri_massive}}
\end{figure}

\subsubsection{Resistive case}

\subsubsection{Qualitative interpretation} 
In order to make sense of the above results, we return to ideal MHD. 
We consider regions close to the disk midplane ($z\sim 0$), where
self-gravity is expected to be most important. For this discussion we
will ignore stratification and set $d/dz\to \imgi k_z$ where $k_z$ is
a vertical wavenumber. The governing equations are then

\begin{align}
  &  0= v_A^2k^2\dvx + \imgi\sigma\left(\imgi\sigma \dvx - 2\Omega\dvy + \imgi k_x \w\right),\label{simp1}\\
  &  0= v_A^2k_z^2\left(\dvy + \frac{\imgi S
  }{\sigma}\dvx\right) + \imgi\sigma\left(\imgi\sigma\dvy +
  \frac{\kappa^2}{2\Omega}\dvx\right),\label{simp2}\\
  & 0 = -k_z^2\w + \frac{\sigma^2}{c_s^2}W + \sigma k_x \dvx, \label{simp3}\\
  & 0 = k^2\dphi + \frac{\Omega^2}{c_s^2Q}W,\label{simp4}
\end{align}
where $k^2 = k_z^2 + k_x^2$.

@@

We now include self-gravity. In general, the MRI will have some
density perturbation so $W^{(0)}\neq0$. The Poisson equation implies
an associated potential perturbation,    
\begin{align} 
  \dphi = -\frac{\Omega^2}{c_s^2 Q k^2} W^{(0)}.
\end{align}
Physically, we expect $k^2\geq0$, so that a positive (negative) density
perturbation causes a negative (positive) potential perturbation. We
then regard $\dphi$ to alter the Cowling solution, so
that $\dvx^{0} \to \dvx^{0} + \dvx^{(1)}$ and similarly for $\dvy$ and
$W$. Then 
\begin{align}
  &   k_x\sigma \dphi = v_A^2k^2\dvx^{(1)} + \imgi\sigma\left[\imgi\sigma
  \dvx^{(1)} - 2\Omega\dvy^{(1)} + \imgi k_x W^{(1)}\right], \label{simp_pert1}\\ 
  &  0= v_A^2k_z^2\left[\dvy^{(1)} + \frac{\imgi S
    }{\sigma}\dvx^{(1)}\right] + \imgi\sigma\left[\imgi\sigma\dvy^{(1)} +
  \frac{\kappa^2}{2\Omega}\dvx^{(1)}\right],\label{simp_pert2}\\
  & k_z^2\dphi  = \left(\frac{\sigma^2}{c_s^2}-k_z^2\right)W^{(1)} +
  \sigma k_x \dvx^{(1)}.\label{simp_pert3} 
\end{align}
Now, if the perturbations to the magnetic field remain
unchanged, i.e. the mode remains close to the standard MRI, then
$\dvx^{(1)} \sim 0$ and $\dvy^{(1)}\sim0$, so Eq. \ref{simp_pert2} is
satisfied. Eq. \ref{simp_pert1} then require $\dphi +
W^{(1)}\sim0$. This is compatible with Eq. \ref{simp_pert3} if 
\begin{align}
  \left|k_z^2\right| \gg \left|\frac{\sigma^2}{c_s^2}\right|. \label{cond}
\end{align}
For the ideal MRI,  we assume $k_z\sim \Omega/v_A$. Then
$|\sigma^2/c_s^2k_z^2|\sim |\sigma^2/\Omega^2\beta|\ll1$ because
$|\sigma|\lesssim \Omega$ and we are considering $\beta\gtrsim
10$. Thus Eq. \ref{cond} is generally satisfied.

The above assumptions imply
\begin{align}
  W^{(1)} \sim \frac{\Omega^2}{c_s^2 Q k^2} W^{(0)},\label{feedback}
\end{align} 
which indicates a non-zero density perturbation due to the
MRI can be amplified by self-gravity. Now, for $k_xH\sim 1$ we have
$|k_z^2/k_x^2|\sim \beta/f\gg1$ because $f=O(1)$ and $\beta\gtrsim10$
for the cases considered above. Then
\begin{align}
  \left|\frac{W^{(1)}}{W^{(0)}}\right| \sim \frac{1}{Q\beta}, 
\end{align}
 suggesting stronger amplification of the density field by 
 self-gravity with increasing field strength (decreasing
 $\beta$).

 For weak fields, the MRI has small vertical wavelength,
 implying rapid density variations with height. The magnitude of the
 corresponding gravitational potential is then expected to be small
 because of internal cancellations, since the potential is essentially
 the integral of the density field.

% not expect significant potential perturbations

%Conversely, for strong fields, only a single MRI wavelength occupies
%the fluid column.

The above arguments also apply to the resistive
limit. Eq. \ref{simp3}---\ref{simp4} are unchanged, while resistive
terms appearing in Eq. \ref{simp1}---\ref{simp2} only involve the
potential perturbation through $\w$. For the resistive MRI we assume
$k_z\sim v_A/\eta$ and $|\sigma|\sim v_A^2/\eta = \Lambda\Omega$
\citep{sano99}. Then $|\sigma^2/c_s^2k_z^2|\sim 1/\beta \ll 1$ so
Eq. \ref{cond} is satisfied. Noting that $k_z^2\sim
\Lambda^2\Omega^2\beta/c_s^2$, the feedback equation becomes
\begin{align}
  \left|\frac{W^{(1)}}{W^{(0)}}\right| \sim
  \frac{1}{Q\left(f^2\hat{k}_x^2 + \beta\Lambda^2\right)},
\end{align}
so increasing the resistivity (decreasing $\Lambda$) should enhance
density perturbations.
